---
subtitle: Data manipulation and tidying data
date: "23rd October, 2020"
output: 
  ioslides_presentation: 
    smaller: no
    logo: mvtec-cover-statistical-programming-4x3.png
    css: style.css
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(tibble.print_min = 5, tibble.print_max = 5)
```

## Today's session

```{r, echo=FALSE, results='asis'}
cat(readr::read_lines("statprog_02_content.md"), sep='\n')
```

# `dplyr`: a grammar of data manipulation

## Datasets {.smaller}

```{r, warning=FALSE, echo=TRUE, results='hide', message=FALSE}
library(nycflights13)
data(package = 'nycflights13')
```

```
Data sets in package ‘nycflights13’:

airlines                      Airline names.
airports                      Airport metadata
flights                       Flights data
planes                        Plane metadata.
weather                       Hourly weather data
```

## Dataset: `flights` {.smaller}

```{r, warning=FALSE, echo=TRUE}
flights
```

## `dplyr` summary

* `filter()` picks cases based on their values.
* `arrange()` changes the ordering of the rows.
* `select()` picks variables based on their names.
* `mutate()` adds new variables that are functions of existing variables.
* `summarise()` reduces multiple values down to a single summary.
* `group_by()` allows to perform any operation “by group”.

<br><br>

Cheat sheet [pdf](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)

## Row filtering (`filter()`)  {.smaller}

* <div class="green2">Selecting all flights on January 1st</div>(`month == 1` and `day == 1`)

```{r}
filter(flights, month == 1, day == 1)
```

## Row filtering (`filter()`) {.smaller}

* <div class="green2">Selecting all flights on January 1st</div>(`month == 1` and `day == 1`)
* <div class="green2">Selecting all flights from six first years of the year</div>(`month <= 6`)
* <div class="green2">Selecting all flights departing with no more that 5 minutes from scheduled departure time</div>(`-5 <= dep_delay` and `dep_delay <= 5` ) or (`abs(dep_delay) <= 5`) or (`between(dep_delay, -5, 5)`)

Exercise.

* <div class="blue2">Select flights flying to "IAH" or "HOU"</div>
* <div class="blue2">Departed in summer (July, August and September)</div>

## More row-selection functions {.smaller}

* `slice()`. Select rows by position (helpers `slice_head()`, `slice_tail()`, `slice_min()`, `slice_max()`, `slice_sample()`)
* `distinct()`. Select distinct observations given certain variables
* `sample_n()`, `sample_frac()`. Random selection of rows

## Ordering observations (`arrange()`) {.smaller}

* Order flights by year, month, day and delay

```{r}
arrange(flights, year, month, day, dep_delay)
```

## Ordering observations (`arrange()`) {.smaller}

* Order flights by year, month, day and __decreasing__ delay

```{r}
arrange(flights, year, month, day, desc(dep_delay))
```

## Selecting variables (`select()`) {.smaller}

* Select year, month, day and flight number

```{r}
select(flights, year, month, day, flight)
```

## `select()` helpers {.smaller}

Special functions can be used to facilitate variable selection.

* Matching names: `starts_with()`, `ends_with()`, `contains()`, `matches()`, `num_range()`. 
* From vector of names: `all_of()`, `any_of()`. 
* Using a function: `where()`

```{r}
select(flights, starts_with('dep_'), contains('arr_'))
```

## Transform variables (`mutate()`) {.smaller}

* Calculating average flight speed (km/h)

```{r}
mutate(flights,
       distance_km = distance * 1.60934,
       air_time_h = air_time * 60,
       speed_km_h = distance_km / air_time_h)
```

## Summarising variables (`summarise()`) {.smaller}

* Calculate the average delays in departure and arrival

```{r}
summarise(flights,
          average_dep_delay = mean(dep_delay, na.rm=TRUE),
          average_arr_delay = mean(arr_delay, na.rm=TRUE))
```

## Working by groups (`group_by()`) {.smaller}

* Stratify `flight` table by day

```{r}
flights_by_day = group_by(flights, year, month, day)
```

* Calculate the average delays in departure and arrival __by day__

```{r, include=FALSE}
options(dplyr.summarise.inform=FALSE)
```

```{r}
summarise(flights_by_day, 
          average_dep_delay = mean(dep_delay, na.rm=TRUE),
          average_arr_delay = mean(arr_delay, na.rm=TRUE))
```

## The pipe operator (`%>%`) {.smaller}

* Summarise the `arr_delay` variable for each month (giving the number of flights and the average delay), for those flights that had a positive `dep_delay`.

```{r, eval=FALSE}
summarise(group_by(filter(flights, dep_delay > 0), month),
          average_arr_delay = mean(arr_delay, na.rm = TRUE))
```

```{r, eval=FALSE}
flights_filtered = filter(flights, dep_delay > 0)
flights_grouped = group_by(flights_filtered, month)
summarise(flights_grouped,
          average_arr_delay = mean(arr_delay, na.rm = TRUE))
```

    * The `%>%` approach:

```{r, eval=FALSE}
flights %>%
  filter(dep_delay > 0) %>%
  group_by(month) %>%
  summarise(
    average_arr_delay = mean(arr_delay, na.rm = TRUE))
```

## Other usefull `dplyr` functions

* Mutating joins: `inner_join()`, `left_join()`, `right_join`, ...
* `expand()`, `expand_grid()`. Create a tibble from all combinations of inputs.
* `bind_rows()`, `bind_cols()`. Efficiently bind multiple data frames by row and column.

# Descriptive statistics: Univariate analysis

## Summarising a categorical variable {.smaller}

* <div class='blue2'>To understand the distribution of a categorical variable, we need the frequencies of its categories: either absolute or relative.</div>
* How is the distribution of `origin`?

```{r}
flights %>%
  group_by(origin) %>%
  summarise(n = n()) %>%
  ungroup() %>%  # <- avoidable if .groups = 'drop' is passed to summarise()
  mutate(p = prop.table(n))
```

## Summarising a categorical variable {.smaller}

* <div class='blue2'>To understand the distribution of a categorical variable, we need the frequencies of its categories: either absolute or relative.</div>
* How is the distribution of `origin`?

```{r}
flights %>%
  count(origin) %>%
  mutate(p = prop.table(n))
```

## Summarising a numerical variable {.smaller}

* <div class='blue2'>To understand the distribution of a numerical variable, the most common aspect are: the location and the variability of its values. The skewness can be of interest too.</div>
* Location. 
    * Mean (`mean()`)
    * Quartiles, deciles, quantiles (`quantile()`)
    * Median (`median()`)
* Spread. 
    * Range (`diff(range())`)
    * Standard deviation (`sd()`)
    * Coefficient of variation (`sd()/mean()`)
    * Interquartile range (`IQR()`)

## Summarising a numerical variable {.smaller}

* How is the distribution of `dep_delay`?

```{r}
summarise_dep_delay = function(.data) 
  summarise(.data,
            mean_ = mean(dep_delay, na.rm = TRUE),
            median_ = median(dep_delay, na.rm = TRUE),
            q1_ = quantile(dep_delay, 0.25, na.rm = TRUE),
            q3_ = quantile(dep_delay, 0.75, na.rm = TRUE),
            iqr_ = IQR(dep_delay, na.rm = TRUE),
            min_ = min(dep_delay, na.rm = TRUE),
            max_ = max(dep_delay, na.rm = TRUE) )
flights %>% summarise_dep_delay()
```

* Highly right-skewed distribution (_mean_ much higher than _median_)
* Distribution centered around zero (_median_ equal to -2)
* 50% of central observations between -5 and 11 ($Q_1$ and $Q_3$)

## Relation between numerical and categorical variables {.smaller}

* How is the distribution of `dep_delay` and `origin`?

```{r}
flights %>%
  group_by(origin) %>%
  summarise_dep_delay()
```

* Distributions of `dep_delay` are similar with respect `origin`.
* Lower variability of `dep_delay` in `LGA` airport.

## Relation between categorical variables {.smaller}

Let's create an `arrival` variable with two values: "delayed" and "on-time".

```{r}
flights = flights %>% 
  mutate(arrival = if_else(arr_delay > 0, 'delayed', 'on-time'))
```

* How is the distribution of `origin` and `on.time`

```{r}
tab = flights %>%
  with(table(arrival, origin))
tab %>% addmargins()
```

## Relation between categorical variables.  {.smaller}

__The marginals__

* Row relative frequencies

```{r}
tab %>% prop.table(margin = 1)
```

* Column relative frequencies

```{r}
tab %>% prop.table(margin = 2)
```

## Relation between numerical variables {.smaller}

* How is the distribution of `dep_delay` and `arr_delay`?

```{r}
flights %>%
  summarise(
    m_dep_delay = median(dep_delay, na.rm = TRUE),
    iqr_dep_delay = IQR(dep_delay, na.rm = TRUE),
    m_arr_delay = median(arr_delay, na.rm = TRUE),
    iqr_arr_delay = IQR(arr_delay, na.rm = TRUE),
    s_dep_arr_delay = cor(dep_delay, arr_delay, use = 'complete.obs')
  )
```

* Higher variability in the arrival delay.
* Strong association between `dep_delay` and `arr_delay`.

## About correlation

* Correlation can be easily misinterpreted.
  * To measure the effect, we need to be free of confounding. [Causal inference course](https://www.bradyneal.com/causal-inference-course)
  * Correlation is not valid for relative data.
      * Variation array for compositional data.
      * Proportionality: A Valid Alternative to Correlation for Relative Data. [Reference](https://pubmed.ncbi.nlm.nih.gov/25775355/)

# Wrangling

## Tidy Data

Tidy data is a standard way of mapping the meaning of a dataset to its structure.

1. Each variable forms a column
2. Each observation forms a row
3. Each type of observational unit forms a table

## `tidyr`: helpers to create tidy data

* "Pivotting". `pivot_longer()` and `pivot_wider()`
* "Rectangling".
* "Nesting".
* Splitting and combining: `separate()` and `unite()`.

<br><br><br><br>

Cheat Sheet [pdf](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf)

## Messy datasets (1) {.smaller} 

__Column headers are values, no variable names__

```{r}
tidyr::relig_income
```

## Messy datasets (2) {.smaller}   

__Multiple variables stored in a column__

```{r}
tidyr::who
```

## Messy datasets (3) {.smaller} 

__Variables are stored in both rows and columns__

```{r, message=FALSE}
library(readr)
read_csv("session-02-presentation/weather.csv")
```

## Dates and time (`lubridate`) (1) {.smaller}

* Parsing dates: `ymd()`, `ydm()`, `myd()`, `mdy()`, `dmy()`, `dym()`

```{r, message=FALSE}
library(lubridate)
x <- c(20090101, "2009-01-02", "2009 01 03", "2009-1-4",
       "2009-1, 5", "Created on 2009 1 6", "200901 !!! 07")
(ddates = ymd(x))
```

* Extracting `year`, `month` or `day`

```{r}
year(ddates)
```

## Dates and time (`lubridate`) (2) {.smaller}

* Calculating time differences

```{r}
diff(ymd(20100101) + years(1:5))
```

* How many weeks?

```{r}
interval(ymd(20101001), ymd(20101101)) / weeks(1)
```


Cheat Sheet [pdf](https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf)

## String (`stringr`)

A cohesive set of functions designed to make working with strings as easy as possible.

* Detect matches: `str_detect()`
* Subset strings: `str_sub()`
* Manage lengths: `str_length()`, `str_trim()`
* Mutate strings: `str_replace()`, `str_to_lower()`, `str_to_title()`
* Join and split: `str_c()`, `str_split()`

Cheat Sheet [pdf](https://raw.githubusercontent.com/rstudio/cheatsheets/master/strings.pdf)

## Categorical variables (`forcats`)

A suite of tools that solve common problems with factors:

* `fct_reoder()`: reorder the levels of a factor according to some function
* `fct_infreq()`: reorder the levels of a factor according to category frequencies
* `fct_relevel()`: reorder the levels of a factor manually
* `fct_lump_min()`, `fct_lump_n()`, `fct_lump_prop()`: collapse the least frequent values

```{r, include=FALSE}
x <- factor(rep(LETTERS[1:9], times = c(40, 10, 5, 27, 1, 1, 1, 1, 1)))
x %>% fct_lump_n(3) %>% table()
```

Cheat Sheet [pdf](https://raw.githubusercontent.com/rstudio/cheatsheets/master/factors.pdf)

# That's all for today

## Next week session

```{r, echo=FALSE, results='asis'}
cat(readr::read_lines("statprog_03_content.md"), sep='\n')
```
